---
title: "Scatter Plots"
output: html_fragment
---


## Scatter Plots

Scatter plots represent the values of two different variables using cartesian coordinates. Scatter plots are often used to display relationships between variables. In
{ggplot2}, the `geom_point()` function is used to produce scatter plots. The data used in the examples comes from the `{palmerpenguins}` package.

---

### A simple scatter plot

The simplest scatter plot will look at the relationship between `body_mass_g` and `bill_length_mm` across all types of penguins. Here, `body_mass_g` is represented on the x-axis, and `bill_length_mm` on the y-axis.

:::: {style="display: flex;"}
::: {style="width: 50%;"}
```{r, eval = FALSE}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point()
```
:::
::: {style="width: 50%;"}
```{r, echo=FALSE, warning=FALSE}
library(ggplot2)
library(palmerpenguins)
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point()
```
:::
::::

---

### Colouring by group

An additional variable can be represented by colouring the points based on the value of a third variable. This is done by setting the `colour` within the aesthetic mapping in `aes()` colours the points. Here, we colour by the `species` of penguin.

:::: {style="display: flex;"}
::: {style="width: 50%;"}
```{r, eval = FALSE}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point()
```
:::
::: {style="width: 50%;"}
```{r, echo=FALSE, warning=FALSE}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point()
```
:::
::::

---

### Adding a fitted line

In scatter plots, it is often desirable to include a best-fitting line to the points. The function `geom_smooth()` adds a line and confidence intervals. By setting the `method` argument to `"lm"`, a straight line is fitted to the data set.

:::: {style="display: flex;"}
::: {style="width: 50%;"}
```{r, eval = FALSE}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x)
```
:::
::: {style="width: 50%;"}
```{r, echo=FALSE, warning=FALSE}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = bill_length_mm)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x)
```
:::
::::

### Final version

When, the data is grouped by species (through setting the colour argument), `geom_smooth()` fits three different best-fitting lines: one for each species. Setting the `se` argument to `FALSE` removes the confidence intervals from the plot. 

`scale_color_brewer()` can be used to set a nicer colour scheme. The `"Dark2"` option is colour-blind friendly. A title, subtitle, caption, and axis titles are set through the `labs()` function. The built-in light {ggplot2} theme is added in here using `theme_light()`.

The legend is moved inside the plotting area using `legend.position` within the `theme()` function. The value of `c(1,0)` places it in the bottom right corner. `legend.justification` determines which part of the legend box is aligned at `c(1,0)`. Here, the bottom right corner of the legend box is aligned with the bottom right corner of the plot area.

:::: {style="display: flex;"}
::: {style="width: 50%;"}
```{r, eval = FALSE}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, se = FALSE) +
  scale_color_brewer("Species", palette = "Dark2") +
  labs(title = "Body Mass and Bill Length of Different Penguin Species",
       subtitle = "Penguins with higher body mass tend to also have longer bills.",
       caption = "Data: {palmerpenguins}",
       x = "Body Mass (g)",
       y = "Bill Length (mm)") +
  theme_light() +
  theme(legend.position = c(1, 0),
        legend.justification = c(1, 0))
```
:::
::: {style="width: 50%;"}
```{r, echo=FALSE, warning=FALSE}
ggplot(data = penguins,
       mapping = aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, se = FALSE) +
  scale_color_brewer("Species", palette = "Dark2") +
  labs(title = "Body Mass and Bill Length of Different Penguin Species",
       subtitle = "Penguins with higher body mass tend to also have longer bills.",
       caption = "Data: {palmerpenguins}",
       x = "Body Mass (g)",
       y = "Bill Length (mm)") +
  theme_light() +
  theme(legend.position = c(1, 0),
        legend.justification = c(1, 0))
```
:::
::::

---



