---
title: "Custom themes"
output: html_fragment
---
  

## Custom themes

Bubble plots are a special type of scatter plot, where the size of the points is mapped to a third variable. In {ggplot2}, the `geom_point()` function is used to produce bubble plots. The data used in the examples comes from the `{palmerpenguins}` package. 

There are 93 elements of You can view them by running `names(theme_minimal())`

Each part of the theme 
* `element_text()`

--- 

### Default theme

:::: {style="display: flex;"}
::: {style="width: 50%;"}
```{r, eval = FALSE}
g <- ggplot(data = penguins, 
       mapping = aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, se = FALSE) +
  scale_colour_discrete(name = "Species") +
  labs(title = "Body Mass and Bill Length of Different Penguin Species", 
       x = "Body Mass (g)", 
       y = "Bill Length (mm)") 
g
```
:::
::: {style="width: 50%;"}
```{r, echo=FALSE, warning=FALSE}
library(ggplot2)
library(palmerpenguins)
g <- ggplot(data = penguins, 
       mapping = aes(x = body_mass_g, y = bill_length_mm, colour = species)) +
  geom_point() +
  geom_smooth(method='lm', formula= y~x, se = FALSE) +
  scale_colour_discrete(name = "Species") +
  labs(title = "Body Mass and Bill Length of Different Penguin Species",
       x = "Body Mass (g)", 
       y = "Bill Length (mm)") 
g
```
:::
::::

---

### Background colours

:::: {style="display: flex;"}
::: {style="width: 50%;"}
```{r, eval = FALSE}
g 
```
:::
::: {style="width: 50%;"}
```{r, echo=FALSE, warning=FALSE}
g
```
:::
::::

---

### Legends

---

### Fonts

The same arguments can be applied to:

Note that some versions of Ubuntu fail to render `"mono"`. Try `"monospace"` instead.

:::: {style="display: flex;"}
::: {style="width: 50%;"}
```{r, eval = FALSE}
g + theme(plot.title = element_text(size = 18, 
                                    colour = "blue", 
                                    face = "bold",
                                    family = "serif",
                                    hjust = 0.5, 
                                    vjust = 1))
```
:::
::: {style="width: 50%;"}
```{r, echo=FALSE, warning=FALSE}
g + theme(plot.title = element_text(size = 18, 
                                    colour = "blue", 
                                    face = "bold",
                                    family = "serif",
                                    hjust = 0.5, 
                                    vjust = 1))
```
:::
::::

You can also use 

---

### Gridlines

---
